---
title: "Computational Musicology Portfolio"
author: "Mick"
date: "2024-02-12"
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    theme:
      version: 4
      bootswatch: cerulean
    vertical_layout: scroll
---

```{=html}
<style>
.dataTables_scrollBody {
    height:540px !important;
}
.chart-stage-flex {
    overflow:auto !important;
}
</style>
```

```{r, echo=FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

```{r}
  library(tidyverse)
  library(spotifyr)
  library(dplyr)
  library(ggplot2)
  library(flexdashboard)
  library(plotly)
  library(compmus)
```

```{r}
Top50Netherlands <- get_playlist_audio_features("", "37i9dQZEVXbKCF6dqVpDkS")
Top50USA <- get_playlist_audio_features("", "37i9dQZEVXbLRQDuF5jeBp")
Top50Brazil <- get_playlist_audio_features("", "37i9dQZEVXbMXbN3EUUhlg")
Top50Australia <- get_playlist_audio_features("", "37i9dQZEVXbJPcfkRz0wJ0")
Top50Philippines <- get_playlist_audio_features("", "37i9dQZEVXbNBz9cRCSFkY")
Top50Morocco <- get_playlist_audio_features("", "37i9dQZEVXbJU9eQpX8gPT")
Top50UK <- get_playlist_audio_features("", "37i9dQZEVXbLnolsZ8PSNw")
```

# Corpus

## Column

### Introduction

The corpus that will be used for this project is the top 50 of multiple countries on Spotify that has an available top 50. This is an interesting corpus because it can show the differences and similarities between popular music in many different places across the world. The comparison points can be either countries, regions, or continents. This flexibility is another advantage of the corpus. The limitations of this corpus are the fact that recent releases or events can influence music worldwide and can push music to the top of many top 50 lists while not saying much about the differences between places (such as the recent release of Kanye West having multiple songs in multiple different top 50 lists). Another limitation is the fact that multiple countries might not use Spotify as their main source of music, so this could influence regional trends in a way that makes them not representative of the population of the country as a whole. A final weakness is the fact that this corpus changes daily, so it can be hard to look into individual tracks since they will probably be gone after a certain time. Finally, it is difficult to pick specific interesting tracks in this corpus due to the sheer number of tracks (3600) in it. But some that may be interesting at first glance are CARNIVAL (Kanye West), Cruel Summer (Taylor Swift), Unwritten (Natasha Bedingfield) and I Wanna Be Yours (Arctic Monkeys). This Kanye West song and many others of his (very) recent album are interesting because they can be found in nearly every top 50 list. The other three songs are mostly interesting because they are older songs that are still found in multiple top 50 lists across the corpus.

# Outlier analysis {data-navmenu=Analysis data-orientation=rows}

## Row 

### Release date of the songs

#### Plot A
This plot shows the release date of all songs in the top 50 of all the different regions, I expected to see mostly very new songs with maybe a few outliers. However, surprisingly a lot of songs are from before 2022, some are even as old as before 2000! This plot also shows some regional differences, the Philippines and the UK seem to enjoy older songs the most, with multiple songs from between 1998 to 2014 making their top 50s, the USA seem to enjoy songs from between 2014 to 2022. Morocco, the Netherlands, Brazil and Australia only seem to like newer songs with a handful of exceptions.

#### Plot B
This plot shows the relation between energy and valence in music from all the different top 50s. It shows that energy and valence are correlated, but not as much as you might expect. It also shows that nearly every popular song has an energy value of 0.4 or higher, those that do not always have a valence level of ~0.5 or lower. This seems to mean that while every "emotional" value is represented in the different top 50s, there does seem to be a minimal amount of energy needed for a song to become popular. A few regional differences can be observed, first of all, Brazil and the UK seem to enjoy more energetic and happier music than average, while the Philippines is the opposite. The other countries all seem to have a similar distribution in energy and valence.

## Row  {.tabset .tabset-fade style="min-height:600px"}

### Plot A {data-width="650"}
```{r}
Top50s <-
  bind_rows(
    mutate(Top50Netherlands, category = "Top50Netherlands"),
    mutate(Top50USA, category = "Top50USA"),
    mutate(Top50Brazil, category = "Top50Brazil"),
    mutate(Top50Australia, category = "Top50Australia"),
    mutate(Top50Philippines, category = "Top50Philippines"),
    mutate(Top50Morocco, category = "Top50Morocco"),
    mutate(Top50UK, category = "Top50UK")
  )
```

```{r}
country_colors <- c("Top50USA" = "#b22222", "Top50Australia" = "#006400", "Top50Brazil" = "#ffd700", 
                    "Top50Netherlands" = "#ff4500", "Top50Philippines" = "#0000ff", "Top50Morocco" = "#8b0000", "Top50UK" = "#000080")
```

```{r}
# Convert release_date column to Date object
Top50s$track.album.release_date <- as.Date(Top50s$track.album.release_date)

# Create the ggplot and save it as a variable
my_plot <- Top50s %>% 
  ggplot(aes(x = track.album.release_date, y = track.popularity, label=track.name)) + 
  geom_jitter(aes(color = category)) + 
  geom_smooth() + 
  scale_color_manual(values = country_colors)  +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y")

# Render the ggplot using ggplotly
ggplotly(my_plot)
```

### Plot B
```{r}
# Create the ggplot and save it as a variable
my_plot <- Top50s %>% 
  ggplot(aes(x = energy, y = valence, label=track.name)) + 
  geom_jitter(aes(color = category)) + 
  geom_smooth() + 
  scale_color_manual(values = country_colors)

# Render the ggplot using ggplotly
ggplotly(my_plot)
```

# Regional differences {data-navmenu=Analysis data-orientation=rows}

#### Description
Some text discussing all the different plots.

## Column {.tabset .tabset-fade}

```{r}
# Function to capitalize the first letter of a string
capitalize_first <- function(x) {
  paste(tools::toTitleCase(x), collapse = " ")
}

# Function to generate violin plot for any feature
plot_feature <- function(data, feature) {
  feature_str <- capitalize_first(deparse(substitute(feature)))  # Capitalize the first letter of the feature
  
  # Calculate the average feature value for each country
  avg_feature <- data %>%
    group_by(category) %>%
    summarise(avg_value = mean({{ feature }})) %>%
    arrange(desc(avg_value))  # Order by average feature value from high to low
  
  # Create the violin plot
  ggplot(data, aes(x = reorder(category, -{{ feature }}), y = {{ feature }}, fill = category)) +
    geom_violin(trim = FALSE) +
    geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white", color = "black", alpha = 0.5) +
    scale_fill_manual(values = country_colors) +
    theme_minimal() +
    labs(
      title = paste("Distribution of", feature_str, "Levels Across Playlists"),
      x = "Playlist",
      y = paste(feature_str, "Level"),
      fill = "Country"
    ) +
    theme(
      legend.position = "none",
      plot.title = element_text(hjust = 0.5),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
}


```

### Acousticness
```{r}
# plot acousticness
plot_feature(Top50s, acousticness)
```

### Danceability
```{r}
# plot danceability
plot_feature(Top50s, danceability)
```

### Energy
```{r}
# plot energy
plot_feature(Top50s, energy)
```

### Liveness
```{r}
# plot liveness
plot_feature(Top50s, liveness)
```

### Loudness
```{r}
# plot loudness
plot_feature(Top50s, loudness)
```

### Speechiness
```{r}
# plot speechiness
plot_feature(Top50s, speechiness)
```

### Tempo
```{r}
# plot tempo
plot_feature(Top50s, tempo)
```

### Valence
```{r}
# plot valence
plot_feature(Top50s, valence)
```

# Individual songs {data-navmenu=Analysis data-orientation=rows}

## Row

### Description

This is a chroma graph of an outlier in the dataset: "Cruel Summer" by Taylor Swift, which is one of the oldest songs which is found in multiple top 50 lists.

## Column

### Chroma analysis of an outlier

```{r, fig.width=15, fig.height=4}
iris <-
  get_tidy_audio_analysis("6Qyc6fS4DsZjB2mRW9DsQs") %>% 
  select(segments) %>% 
  unnest(segments) %>% 
  select(start, duration, pitches)

iris %>% 
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) %>% 
  compmus_gather_chroma() %>% 
  ggplot(
    aes(
      x = start + duration / 2,
      width = duration,
      y = pitch_class,
      fill = value
    )
  ) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  theme_minimal() +
  scale_fill_viridis_c()
```


## Row {style="min-height:600px"}

### Structure similarity matrix of an outlier

```{r}
what_was_i_made_for <-
  get_tidy_audio_analysis("6wf7Yu7cxBSPrRlWeSeK0Q") %>%
  compmus_align(bars, segments) %>%
  select(bars) %>%
  unnest(bars) %>%
  mutate(
    pitches =
      map(segments,
        compmus_summarise, pitches,
        method = "acentre", norm = "manhattan"
      )
  ) %>%
  mutate(
    timbre =
      map(segments,
        compmus_summarise, timbre,
        method = "mean"
      )
  )
bind_rows(
  what_was_i_made_for %>%
    compmus_self_similarity(pitches, "aitchison") %>%
    mutate(d = d / max(d), type = "Chroma"),
  what_was_i_made_for %>% 
    compmus_self_similarity(timbre, "euclidean") %>% 
    mutate(d = d / max(d), type = "Timbre")
) %>%
  mutate() %>% 
  ggplot(
    aes(
      x = xstart + xduration / 2,
      width = xduration,
      y = ystart + yduration / 2,
      height = yduration,
      fill = d
    )
  ) +
  geom_tile() +
  coord_fixed() +
  facet_wrap(~type) +
  scale_fill_viridis_c(option = "E", guide = "none") +
  theme_classic() + 
  labs(x = "", y = "")
```

### Description

These are two self similarity matrices of  "Cruel Summer" by Taylor Swift, these similarity matrices show both Chroma and Timbre.

## Row {style="min-height:600px"}

### Description
This chordogram shows the chords in "End of Beginning" by Djo, this song shows six clear sections. This is a typical feature of many popular songs in the corpus, because the very similar sections are the chorus returning three times which seems to be a key feature in creating a popular songs.

### Chordogram

```{r}
circshift <- function(v, n) {
  if (n == 0) v else c(tail(v, n), head(v, -n))
}

#      C     C#    D     Eb    E     F     F#    G     Ab    A     Bb    B
major_chord <-
  c(   1,    0,    0,    0,    1,    0,    0,    1,    0,    0,    0,    0)
minor_chord <-
  c(   1,    0,    0,    1,    0,    0,    0,    1,    0,    0,    0,    0)
seventh_chord <-
  c(   1,    0,    0,    0,    1,    0,    0,    1,    0,    0,    1,    0)

major_key <-
  c(6.35, 2.23, 3.48, 2.33, 4.38, 4.09, 2.52, 5.19, 2.39, 3.66, 2.29, 2.88)
minor_key <-
  c(6.33, 2.68, 3.52, 5.38, 2.60, 3.53, 2.54, 4.75, 3.98, 2.69, 3.34, 3.17)

chord_templates <-
  tribble(
    ~name, ~template,
    "Gb:7", circshift(seventh_chord, 6),
    "Gb:maj", circshift(major_chord, 6),
    "Bb:min", circshift(minor_chord, 10),
    "Db:maj", circshift(major_chord, 1),
    "F:min", circshift(minor_chord, 5),
    "Ab:7", circshift(seventh_chord, 8),
    "Ab:maj", circshift(major_chord, 8),
    "C:min", circshift(minor_chord, 0),
    "Eb:7", circshift(seventh_chord, 3),
    "Eb:maj", circshift(major_chord, 3),
    "G:min", circshift(minor_chord, 7),
    "Bb:7", circshift(seventh_chord, 10),
    "Bb:maj", circshift(major_chord, 10),
    "D:min", circshift(minor_chord, 2),
    "F:7", circshift(seventh_chord, 5),
    "F:maj", circshift(major_chord, 5),
    "A:min", circshift(minor_chord, 9),
    "C:7", circshift(seventh_chord, 0),
    "C:maj", circshift(major_chord, 0),
    "E:min", circshift(minor_chord, 4),
    "G:7", circshift(seventh_chord, 7),
    "G:maj", circshift(major_chord, 7),
    "B:min", circshift(minor_chord, 11),
    "D:7", circshift(seventh_chord, 2),
    "D:maj", circshift(major_chord, 2),
    "F#:min", circshift(minor_chord, 6),
    "A:7", circshift(seventh_chord, 9),
    "A:maj", circshift(major_chord, 9),
    "C#:min", circshift(minor_chord, 1),
    "E:7", circshift(seventh_chord, 4),
    "E:maj", circshift(major_chord, 4),
    "G#:min", circshift(minor_chord, 8),
    "B:7", circshift(seventh_chord, 11),
    "B:maj", circshift(major_chord, 11),
    "D#:min", circshift(minor_chord, 3)
  )

key_templates <-
  tribble(
    ~name, ~template,
    "Gb:maj", circshift(major_key, 6),
    "Bb:min", circshift(minor_key, 10),
    "Db:maj", circshift(major_key, 1),
    "F:min", circshift(minor_key, 5),
    "Ab:maj", circshift(major_key, 8),
    "C:min", circshift(minor_key, 0),
    "Eb:maj", circshift(major_key, 3),
    "G:min", circshift(minor_key, 7),
    "Bb:maj", circshift(major_key, 10),
    "D:min", circshift(minor_key, 2),
    "F:maj", circshift(major_key, 5),
    "A:min", circshift(minor_key, 9),
    "C:maj", circshift(major_key, 0),
    "E:min", circshift(minor_key, 4),
    "G:maj", circshift(major_key, 7),
    "B:min", circshift(minor_key, 11),
    "D:maj", circshift(major_key, 2),
    "F#:min", circshift(minor_key, 6),
    "A:maj", circshift(major_key, 9),
    "C#:min", circshift(minor_key, 1),
    "E:maj", circshift(major_key, 4),
    "G#:min", circshift(minor_key, 8),
    "B:maj", circshift(major_key, 11),
    "D#:min", circshift(minor_key, 3)
  )

end_of_beginnig <-
  get_tidy_audio_analysis("3qhlB30KknSejmIvZZLjOD") %>%
  compmus_align(sections, segments) %>%
  select(sections) %>%
  unnest(sections) %>%
  mutate(
    pitches =
      map(segments,
        compmus_summarise, pitches,
        method = "mean", norm = "manhattan"
      )
  )

end_of_beginnig %>%
  compmus_match_pitch_template(
    key_templates,         # Change to chord_templates if descired
    method = "euclidean",  # Try different distance metrics
    norm = "manhattan"     # Try different norms
  ) %>%
  ggplot(
    aes(x = start + duration / 2, width = duration, y = name, fill = d)
  ) +
  geom_tile() +
  scale_fill_viridis_c(guide = "none") +
  theme_minimal() +
  labs(x = "Time (s)", y = "")
```
